---
title: "Physical Activity Classification Using Machine Learning"
author: "Ari Kerstein"
date: "September 10, 2016"
output: 
  html_document: 
    keep_md: yes
---
# Summary
This analysis uses data on Human Activity Recognition (HAR) to determine how well study participants performed dumbbell exercises. The data in question--for which more information can be found [here](http://groupware.les.inf.puc-rio.br/har#weight_lifting_exercises)--indicates various metrics about the performance of the exercises along with a $classe$ variable denoting the manner in which the exercises were performed. This analysis uses machine learning algorithms to predict the manner of exercise given the performance metrics available. Prediction accuracy of 90% and 95% on the training and validations sets, respectively, were obtained using a random forest model. The model ultimately scored 90% on the testing set. 
  
# The Data
The training data can be downloaded from the Internet [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv), while the testing data can be found [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv). The data are loaded to R using the following code, assuming the downloaded files are in the user's R working directory.
```{r,warning=FALSE,message=FALSE}
library(caret); library(rpart);library(randomForest)
training<-read.csv("./pml-training.csv")
testing<-read.csv("./pml-testing.csv")
```
The training data has the following properties. 
```{r}
str(training)
```
Two observations from looking at the data: first, there are subject and row indicator variables in columns 1 to 7 that will not be needed for machine learning purposes; second, there are many columns for which a majority of the values are missing (either $NA$ or blank). These columns appear to be those that contain certain statistical measures (e.g. kurtosis, avg). The code below transforms the training data for analysis, removing the first 7 columns and the columns containing mostly missing values. 
```{r}
training<-training[,-c(1:7)] ##Remove first 7 columns
toMatch<-c("^avg","^std","^kurtosis","^skewness","^max","^min",
           "^amplitude","^var")
grep<-unique(grep(paste(toMatch,collapse="|"), 
            names(training)))
training<-training[,-grep] ##Remove columns with missing values
```
Doing this results in a very usable data set for predictive analysis, notably because there are now no missing values for the predictor variables. 
```{r}
dim(training)
sum(complete.cases(training))
```
  
# Predictive Analysis
The first step for predictive analysis is to split the training data into a training set and a validation set, using tools from the **caret** package in R. 
```{r}
set.seed(1234)
inTrain<-createDataPartition(y=training$classe,p=.7,list=FALSE)
training<-training[inTrain,]
validate<-training[-inTrain,]
```
There are 52 predictor variables to help classify the manner of exercise. The strategy employed here will be to run a 10-fold cross validated rpart model using all 52 predictors.
```{r}
set.seed(4567)
tree1<-train(classe~.,data=training,method="rpart",trControl=trainControl(method="cv"))
tree1
```
Unfortunately, the accuracy of this model will likely be no better than 50% on out of sample data. But, the model can tell us which predictors it used to split the nodes of the tree. 
```{r}
tree1$finalModel
```
To enhance the accuracy of this first rpart model, we will use a random forest model and incorporate only the predictor variables that made it into the rpart model. This is done as a proxy for feature selection and as a way to save on computing resources needed to run forests or boosting models on the entire training set. 
```{r}
set.seed(789)
rf1<-train(classe~roll_belt+pitch_forearm+magnet_dumbbell_y+roll_forearm,data=training,method="rf",trControl=trainControl(method="cv"))
rf1
```
This random forest model does much better than the rpart model in terms of accuracy. We can now estimate the out of sample accuracy of the model by predicting results on the validation set. 
```{r}
predrf1<-predict(rf1,newdata=validate)
CMrf1<-confusionMatrix(predrf1,reference=validate$classe)
CMrf1
```
The RF model actually scores higher accuracy on the validation set than it did on the training set. This could indicate the potential for overfitting, but the cross-validation we used is meant to limit this possibility. Given, though, that out of sample results are generally less accurate than those from the training set, it would be safe to estimate that the out of sample accuracy for this model is at or slightly under 90%. 
  
# Conclusion
Using a 10-fold cross validated random forest model on 4 predictors in the training data set, we were able to generate predictions that were about 90% accurate on the training data set and 95% accurate on the validation set. Despite the enhanced performance on the validation set, we would expect a similar or lower out of sample performance on the testing set. Ultimately, outside the scope of this particular write-up, the model scored 90% on the testing set, in line with expectations.  